Assessment of forestry road network projection methods using the “roads” package in R

### 2022-10-05 analysis/paper/paper.Rmd
make figs show dlcp results too

# read in example area data
exRoads <- read_sf(here(data_path_drvd, "testing_ex_roads.gpkg"))
tsb <- read_sf(here(data_path_drvd, "testing_tsb.gpkg"))
cutblocks <- read_sf(here(data_path_drvd, "testing_cutblocks.gpkg"))
tsaCost <- raster(here(data_path_drvd, "testing_cost.tif"))
roads <- read_sf(here(data_path_drvd, "testing_obs_roads.gpkg"))

sampleDens <- c(low,high,low,high,low)
sampleType <- c("regular","regular","random","random","centroid")
paramTable <- tibble(sampleType, sampleDens,
                     runTime = vector("list", length(sampleDens)),
                     output = vector("list", length(sampleDens)),
                     roadDisturbance = vector("list", length(sampleDens)),
                     roadDensity = vector("list", length(sampleDens)),
                     roadPresence = vector("list", length(sampleDens)),
                     distanceToRoad = vector("list", length(sampleDens)),
                     forestryDisturbance = vector("list", length(sampleDens))) %>%
  distinct()

# re-run projections for just the small area with different parameters
mstProj <- projectAll(tsbs = tsb, paramTable = paramTable,
                      costSurface = tsaCost,
                      cutblocks = cutblocks,
                      existingRoads = exRoads,
                      fileLocation = paste0(data_path_drvd, "for_fig/mst"),
                      method = "mst")

dlcpProj <- projectAll(tsbs = tsb,
                      paramTable = paramTable %>% filter(sampleType == "regular"),
                      costSurface = tsaCost,
                      cutblocks = cutblocks,
                      existingRoads = exRoads,
                      fileLocation = paste0(data_path_drvd, "for_fig/dlcp"),
                      method = "dlcp")

lcpProj <- projectAll(tsbs = tsb,
                       paramTable = paramTable %>% filter(sampleType == "centroid"),
                       costSurface = tsaCost,
                       cutblocks = cutblocks,
                       existingRoads = exRoads,
                       fileLocation = paste0(data_path_drvd, "for_fig/lcp"),
                       method = "lcp")

@@

Projection accuracy varied widely across the 5 metrics and two scales compared (Figure \@ref(fig:fig3)). Across all projection methods road projections were less accurate within cutblocks than in the whole TSA. Metrics that used a buffer around the roads, forestry and road disturbance footprint, showed closer agreement between the observed and projected roads than the road density and road presence metrics at both scales. For all methods the distance to road metric was underestimated within cutblocks but was similar to the observed across the whole TSA. The road density and road distance to road metrics were the most sensitive to pixel size and were less accurate when the coarse cost raster was used (\@ref(fig:figA1)).

@@

```{r fig3, fig.height=4, fig.width=fig_widths["mid"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods across the entire Revelstoke TSA (“overall”) and within cutblocks (“cutblocks”). Values greater than 0 indicate the projection overestimated the amount of roads and negative values indicate an underestimation."}
# organize data for ggplot
meanTable_long <- meanTable %>%
  pivot_longer(-c(sampleType, sampleDens, areaMean),
               names_to = "metric", values_to = "response")%>%
  mutate(sampleType = ifelse(sampleType == sampleDens, NA_character_, 
                             sampleType),
         sampleDens = paste(sampleType, sampleDens) %>% 
           factor(levels = c("NA centroid",
                             "NA observed",
                             "random low sample density",
                             "regular low sample density",
                             "random high sample density",
                             "regular high sample density",
                             "NA klementQGIS"),
                  labels = c("Centroid", 
                             "Observed",
                             "Random\nlow density",
                             "Regular\nlow density",
                             "Random\nhigh density",
                             "Regular\nhigh density",
                             "Hardy QGIS")),
         metric = factor(metric,
                         levels = c("roadDensityMean", "roadPresenceMean",
                                    "distanceToRoadMean",
                                    "forestryDisturbanceMean",
                                    "roadDisturbanceMean"),
                         labels = c("Road\ndensity", "Road\npresence",
                                    "Distance\nto road",
                                    "Forestry\ndisturbance\nfootprint",
                                    "Road\ndisturbance\nfootprint")))

@@

prop_dif %>% 
  ggplot(aes(x = sampleDens, prop_dif, fill = sampleDens))+
  geom_hline(aes(yintercept = 0), color = "grey")+
  geom_col(position = position_dodge2(preserve = "single",
                                      width = 0.75))+
  facet_grid(areaMean~ metric)+
  theme_classic()+
  scale_fill_manual(values = met.brewer(pal_nm, 8) %>% .[c(8, 1:4, 6)],
                    name = "Sampling\nMethod", guide = "none")+
  theme(text = element_text(size = 10), axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.position = "right",
        axis.text.x=element_text(size=11, angle=50, vjust=1, hjust=1))+
  coord_flip()

# Create bar plot This is the old version keep in case
fig3 <- projected_values %>% group_by(areaMean) %>% nest() %>% ungroup() %>%
  mutate(obs = observed_values %>% group_by(areaMean) %>% nest() %>% pull(data)) %>%
  group_by(areaMean) %>%
  mutate(data = map(data, ~mutate(.x, area = areaMean)),
         plot = map2(data, obs,
                    ~ ggplot(.x, aes(x = sampleDens, response, fill = sampleType))+
                      geom_col(position = position_dodge2(preserve = "single",
                                                         width = 0.75))+
                      geom_hline(aes(yintercept = response),
                                 .y, color = "black", linetype = "dashed")+
                      facet_wrap(~metric, scales = "free_y", ncol = 1)+
                      theme_classic()+
                      scale_fill_manual(values = met.brewer(pal_nm, 4))+
                      theme(text = element_text(size = 15), axis.title.x = element_blank(),
                            axis.title.y = element_blank(),
                            legend.position = "none")+
                      ggtitle(str_to_title(areaMean)))) %>%
  pull(plot) %>%
  {ggarrange(plotlist = .)}

@@

matchData_sum <- matchData %>% ungroup() %>%
  mutate(agreement = factor(agreement,
                            levels = c("False negative", "False positive",
                                       "Agree roaded","Agree roadless",
                                       "Pre-existing roads")),
         sampleDens = paste(sampleType, sampleDens) %>%
           str_replace("1e-06", "low density") %>%
           str_replace("1e-05", "high density") %>%
           str_replace("centroid low density", "centroid") %>%
           str_replace("klementQGIS NA", "klementQGIS") %>%
           factor(levels = c("centroid",
                             "random low density",
                             "regular low density",
                             "random high density",
                             "regular high density",
                             "klementQGIS"),
                  labels = c("Centroid", 
                             "Random\nlow density",
                             "Regular\nlow density",
                             "Random\nhigh density",
                             "Regular\nhigh density",
                             "Hardy QGIS")),
         metric = factor(metric,
                         levels = c("roadPresence",
                                    "forestryDisturbance",
                                    "roadDisturbance"),
                         labels = c("Road presence",
                                    "Forestry disturbance\nfootprint",
                                    "Road disturbance footprint"))) %>%
  group_by(metric, sampleDens, agreement) %>%
  summarise(count = sum(count), .groups = "drop_last") %>%
  mutate(perc = count/sum(count)*100)

stackedBar <- matchData_sum %>%
  ggplot(aes(x = sampleDens, y =perc, fill = agreement))+
  geom_col(position = "stack", width = 0.75)+
  theme_classic()+
  theme(axis.text.x=element_text(color = "black", size=11, angle=50, vjust=1, hjust=1))+
  scale_fill_manual(values = c(met.brewer(pal_nm, 4, direction = -1)[c(1,4,3,2)],
                               "#999999"))+
  theme(text = element_text(size = 15), axis.title.x = element_blank())+
  theme(legend.position="bottom", legend.text = element_text(size = 10),
        legend.title = element_blank(),
        strip.background = element_blank())+
  labs(y = "Percent of Landscape")+
  facet_wrap(~metric)

# waffle stacked bar
# matchData_sum %>%
#   #filter(metric == "Road presence") %>%
#   ggplot(aes(values = perc*3,
#              fill = factor(agreement, levels = levels(agreement))))+
#   geom_waffle(size = 0.25, colour = "white", flip = TRUE, n_rows = 9)+
#   facet_grid(metric~ sampleDens)+
#   scale_fill_manual(values = c(met.brewer(pal_nm, 4, direction = -1)[c(1,4,3,2)],
#                                "#999999"), name = "Agreement")+
#   scale_x_discrete(name = NULL) +
#   scale_y_continuous(labels = function(x) x * 9/3,
#                      expand = c(0,0), name =  "Percent of Landscape") +
#   coord_equal()+
#   theme_minimal() +
#   theme(text = element_text(size = 10),
#         panel.grid = element_blank(),
#         axis.ticks.y = element_line(),
#         legend.position = "bottom") +
#   guides(fill=guide_legend(nrow=2))

@@

perf_tbl %>% 
  pivot_longer(c(sensitivity, precision, F_measure), names_to = "perf_meas",
               values_to = "value") %>% 
  mutate(perf_meas = factor(perf_meas,
                            labels = c("F measure", "Precision", "Sensitivity"))) %>% 
  ggplot(aes(sampleDens, value, fill = sampleDens))+
  geom_col()+
  scale_fill_manual(values = met.brewer(pal_nm, 8) %>% .[c(8, 1:4, 6)],
                    name = "Sampling\nMethod", guide = "none")+
  facet_grid(metric ~ perf_meas)+
  coord_flip()+
  ylim(0, 1)+
  theme_classic()+
  labs(y = "Performance", x = NULL)

@@

```{r figA1, fig.height=4, fig.width=fig_widths["mid"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods across the entire Revelstoke TSA (“overall”) and within cutblocks (“cutblocks”) for the fine resolutiom (100 m) and coarse resolution (1000 m) cost rasters. Values greater than 0 indicate the projection overestimated the metric and negative values indicate an underestimation."}
# organize data for ggplot
meanTable_long <- meanTable %>%
  pivot_longer(-c(sampleType, sampleDens, areaMean, resolution),
               names_to = "metric", values_to = "response")%>%
  mutate(sampleType = ifelse(sampleType == sampleDens, NA_character_, 
                             sampleType),
         sampleDens = paste(sampleType, sampleDens) %>% 
           factor(levels = c("NA centroid",
                             "NA observed",
                             "random low sample density",
                             "regular low sample density",
                             "random high sample density",
                             "regular high sample density",
                             "NA klementQGIS"),
                  labels = c("Centroid", 
                             "Observed",
                             "Random\nlow density",
                             "Regular\nlow density",
                             "Random\nhigh density",
                             "Regular\nhigh density",
                             "Hardy QGIS")),
         metric = factor(metric,
                         levels = c("roadDensityMean", "roadPresenceMean",
                                    "distanceToRoadMean",
                                    "forestryDisturbanceMean",
                                    "roadDisturbanceMean"),
                         labels = c("Road\ndensity", "Road\npresence",
                                    "Distance\nto road",
                                    "Forestry\ndisturbance\nfootprint",
                                    "Road\ndisturbance\nfootprint")))

@@

matchData_sum <- matchData %>% ungroup() %>%
  mutate(agreement = factor(agreement,
                            levels = c("False negative", "False positive",
                                       "Agree roaded","Agree roadless",
                                       "Pre-existing roads")),
         sampleDens = paste(sampleType, sampleDens) %>%
           str_replace("1e-06", "low density") %>%
           str_replace("1e-05", "high density") %>%
           str_replace("centroid low density", "centroid") %>%
           str_replace("klementQGIS NA", "klementQGIS") %>%
           factor(levels = c("centroid",
                             "random low density",
                             "regular low density",
                             "random high density",
                             "regular high density",
                             "klementQGIS"),
                  labels = c("Centroid", 
                             "Random\nlow density",
                             "Regular\nlow density",
                             "Random\nhigh density",
                             "Regular\nhigh density",
                             "Hardy QGIS")),
         metric = factor(metric,
                         levels = c("roadPresence",
                                    "forestryDisturbance",
                                    "roadDisturbance"),
                         labels = c("Road presence",
                                    "Forestry disturbance\nfootprint",
                                    "Road disturbance footprint"))) %>%
  group_by(metric, sampleDens, agreement, resolution) %>%
  summarise(count = sum(count), .groups = "drop_last") %>%
  mutate(perc = count/sum(count)*100)

>>>>

# read in example area data
exRoads <- read_sf(here(data_path_drvd, "..", "testing_ex_roads.gpkg"))
tsb <- read_sf(here(data_path_drvd, "..", "testing_tsb.gpkg"))
cutblocks <- read_sf(here(data_path_drvd, "..", "testing_cutblocks.gpkg"))
tsaCost <- raster(here(data_path_drvd, "..", "testing_cost.tif"))
roads <- read_sf(here(data_path_drvd, "..", "testing_obs_roads.gpkg"))

sampleDens <- c(low,high,low,high,low)
sampleType <- c("regular","regular","random","random","centroid")
paramTable <- tibble(sampleType, sampleDens,
                     method = "mst",
                     runTime = vector("list", length(sampleDens)),
                     output = vector("list", length(sampleDens)),
                     roadDisturbance = vector("list", length(sampleDens)),
                     roadDensity = vector("list", length(sampleDens)),
                     roadPresence = vector("list", length(sampleDens)),
                     distanceToRoad = vector("list", length(sampleDens)),
                     forestryDisturbance = vector("list", length(sampleDens))) %>%
  distinct()

# re-run projections for just the small area with different parameters
mstProj <- projectAll(tsbs = tsb, paramTable = paramTable,
                      costSurface = tsaCost,
                      cutblocks = cutblocks,
                      existingRoads = exRoads,
                      fileLocation = here(data_path_drvd, "..", "for_fig"))

dlcpProj <- projectAll(tsbs = tsb,
                      paramTable = paramTable %>% 
                        filter(sampleType == "regular") %>% 
                        mutate(method = "dlcp"),
                      costSurface = tsaCost,
                      cutblocks = cutblocks,
                      existingRoads = exRoads,
                      fileLocation = here(data_path_drvd, "..", "for_fig"))

lcpProj <- projectAll(tsbs = tsb,
                       paramTable = paramTable %>% 
                        filter(sampleType == "centroid") %>% 
                        mutate(method = "lcp"),
                       costSurface = tsaCost,
                       cutblocks = cutblocks,
                       existingRoads = exRoads,
                       fileLocation = here(data_path_drvd, "..", "for_fig"))

@@

Projection accuracy varied widely across the 5 metrics and two scales compared (Figure \@ref(fig:fig3)). Across all projection methods road projections were less accurate within cutblocks than in the whole TSA. Metrics that used a buffer around the roads, forestry and road disturbance footprint, showed closer agreement between the observed and projected roads than the road density and road presence metrics at both scales. For all methods the distance to road metric was underestimated within cutblocks but was similar to the observed across the whole TSA. The road density and distance to road metrics were the most sensitive to pixel size and were less accurate when the coarse cost raster was used (\@ref(fig:figA1)).

@@

```{r fig3, fig.height=4, fig.width=fig_widths["mid"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods across the entire Revelstoke TSA (“overall”) and within cutblocks (“cutblocks”). Values greater than 0 indicate the projection overestimated the amount of roads and negative values indicate an underestimation."}
# organize data for ggplot
meanTable_long <- meanTable %>% select(-runTime) %>% 
  pivot_longer(-c(sampleType, sampleDens, areaMean, method),
               names_to = "metric", values_to = "response")%>%
  mutate(sampleType = ifelse(sampleType == sampleDens, NA_character_, 
                             sampleType),
         sampleDens = paste(method, sampleType, sampleDens) %>% 
           factor(levels = c("mst NA centroid",
                             "NA NA observed",
                             "mst random low sample density",
                             "mst regular low sample density",
                             "mst random high sample density",
                             "mst regular high sample density",
                             "dlcp regular high sample density",
                             "NA klementQGIS"),
                  labels = c("Centroid", 
                             "Observed",
                             "Random\nlow density",
                             "Regular\nlow density",
                             "Random\nhigh density",
                             "Regular\nhigh density",
                             "DLCP Regular\nhigh density",
                             "Hardy QGIS")),
         metric = factor(metric,
                         levels = c("roadDensityMean", "roadPresenceMean",
                                    "distanceToRoadMean",
                                    "forestryDisturbanceMean",
                                    "roadDisturbanceMean"),
                         labels = c("Road\ndensity", "Road\npresence",
                                    "Distance\nto road",
                                    "Forestry\ndisturbance\nfootprint",
                                    "Road\ndisturbance\nfootprint")))

@@

prop_dif %>% 
  ggplot(aes(x = sampleDens, prop_dif, fill = sampleDens))+
  geom_hline(aes(yintercept = 0), color = "grey")+
  geom_col(position = position_dodge2(preserve = "single",
                                      width = 0.75))+
  facet_grid(areaMean~ metric)+
  theme_classic()+
  scale_fill_manual(values = met.brewer(pal_nm, 8) %>% .[c(8, 1:6)],
                    name = "Sampling\nMethod", guide = "none")+
  theme(text = element_text(size = 10), axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.position = "right",
        axis.text.x=element_text(size=11, angle=50, vjust=1, hjust=1))+
  coord_flip()

@@

matchData_sum <- matchData %>% ungroup() %>%
  mutate(agreement = factor(agreement,
                            levels = c("False negative", "False positive",
                                       "Agree roaded","Agree roadless",
                                       "Pre-existing roads")),
         sampleDens = paste(method, sampleType, sampleDens) %>%
           str_replace("1e-06", "low density") %>%
           str_replace("1e-05", "high density") %>%
           str_replace("centroid low density", "centroid") %>%
           str_replace("klementQGIS NA", "klementQGIS") %>%
           factor(levels = c("mst centroid",
                             "mst random low density",
                             "mst regular low density",
                             "mst random high density",
                             "mst regular high density",
                             "dlcp regular high density",
                             "klementQGIS"),
                  labels = c("Centroid", 
                             "Random\nlow density",
                             "Regular\nlow density",
                             "Random\nhigh density",
                             "Regular\nhigh density",
                             "DLCP Regular\nhigh density",
                             "Hardy QGIS")),
         metric = factor(metric,
                         levels = c("roadPresence",
                                    "forestryDisturbance",
                                    "roadDisturbance"),
                         labels = c("Road presence",
                                    "Forestry disturbance\nfootprint",
                                    "Road disturbance footprint"))) %>%
  group_by(metric, sampleDens, agreement) %>%
  summarise(count = sum(count), .groups = "drop_last") %>%
  mutate(perc = count/sum(count)*100)

@@

perf_tbl %>% 
  pivot_longer(c(sensitivity, precision, F_measure), names_to = "perf_meas",
               values_to = "value") %>% 
  mutate(perf_meas = factor(perf_meas,
                            labels = c("F measure", "Precision", "Sensitivity"))) %>% 
  ggplot(aes(sampleDens, value, fill = sampleDens))+
  geom_col()+
  scale_fill_manual(values = met.brewer(pal_nm, 8) %>% .[c(8, 1:6)],
                    name = "Sampling\nMethod", guide = "none")+
  facet_grid(metric ~ perf_meas)+
  coord_flip()+
  ylim(0, 1)+
  theme_classic()+
  labs(y = "Performance", x = NULL)

@@

```{r figA1, fig.height=4, fig.width=fig_widths["mid"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection methods across the entire Revelstoke TSA (“overall”) and within cutblocks (“cutblocks”) for the fine resolutiom (100 m) and coarse resolution (1000 m) cost rasters. Values greater than 0 indicate the projection overestimated the metric and negative values indicate an underestimation."}
# organize data for ggplot
meanTable_long <- meanTable %>%
  pivot_longer(-c(sampleType, sampleDens, areaMean, resolution),
               names_to = "metric", values_to = "response")%>%
  mutate(sampleType = ifelse(sampleType == sampleDens, NA_character_, 
                             sampleType),
         sampleDens = paste(method, sampleType, sampleDens) %>% 
           factor(levels = c("mst NA centroid",
                             "NA NA observed",
                             "mst random low sample density",
                             "mst regular low sample density",
                             "mst random high sample density",
                             "mst regular high sample density",
                             "dlcp regular high sample density",
                             "NA klementQGIS"),
                  labels = c("Centroid", 
                             "Observed",
                             "Random\nlow density",
                             "Regular\nlow density",
                             "Random\nhigh density",
                             "Regular\nhigh density",
                             "DLCP Regular\nhigh density",
                             "Hardy QGIS")),
         metric = factor(metric,
                         levels = c("roadDensityMean", "roadPresenceMean",
                                    "distanceToRoadMean",
                                    "forestryDisturbanceMean",
                                    "roadDisturbanceMean"),
                         labels = c("Road\ndensity", "Road\npresence",
                                    "Distance\nto road",
                                    "Forestry\ndisturbance\nfootprint",
                                    "Road\ndisturbance\nfootprint")))

@@

matchData_sum <- matchData %>% ungroup() %>%
  mutate(agreement = factor(agreement,
                            levels = c("False negative", "False positive",
                                       "Agree roaded","Agree roadless",
                                       "Pre-existing roads")),
         sampleDens = paste(method, sampleType, sampleDens) %>%
           str_replace("1e-06", "low density") %>%
           str_replace("1e-05", "high density") %>%
           str_replace("centroid low density", "centroid") %>%
           str_replace("klementQGIS NA", "klementQGIS") %>%
           factor(levels = c("mst centroid",
                             "mst random low density",
                             "mst regular low density",
                             "mst random high density",
                             "mst regular high density",
                             "dlcp regular high density",
                             "klementQGIS"),
                  labels = c("Centroid", 
                             "Random\nlow density",
                             "Regular\nlow density",
                             "Random\nhigh density",
                             "Regular\nhigh density",
                             "DLCP Regular\nhigh density",
                             "Hardy QGIS")),
         metric = factor(metric,
                         levels = c("roadPresence",
                                    "forestryDisturbance",
                                    "roadDisturbance"),
                         labels = c("Road presence",
                                    "Forestry disturbance\nfootprint",
                                    "Road disturbance footprint"))) %>%
  group_by(metric, sampleDens, agreement, resolution) %>%
  summarise(count = sum(count), .groups = "drop_last") %>%
  mutate(perc = count/sum(count)*100)
